---
title: "Schedule"
slug: schedule
summary: "HPAM 7660 schedule with readings, due dates, and links to materials."
toc: false
sidebar: false
page-layout: full
---
           
```{r packages-data, include=FALSE} 
library(tidyverse)
library(knitr)
library(kableExtra)
library(lubridate)
library(glue)

schedule <- read_csv("data/schedule.csv", show_col_types = FALSE) 
```

The detailed course schedule for the semester is below. Please note that dates and topics are subject to change and will depend on the pace of the course. Please stay apprised of any changes announced in class. You can find the materials for each course meeting under the “materials” links for that week in the schedule below. You should generally: 
   
- watch the lecture videos (if any) and complete the readings before the assigned class period
- submit any reading discussion questions, problem sets, or other assignments by 1pm on the assigned due date
                
Here's a guide to the schedule: 

- **Materials** ({{< fa book-open-reader >}}): This page contains the readings, slides, and videos (if any) for the topic. Do the readings and watch the videos before coming to class on the indicated date. Slides will typically be posted before each class.
- **Tutorial** ({{< fa laptop-code >}}): A link to the tutorial for that week (if any). **Note that we will complete tutorials in class - these are NOT homework. 
- **Assignment** ({{< fa pen-ruler >}}): This page contains the instructions for each assignment. **Assignments are due by 1:00 PM on the day they're listed.** 

 
```{r build-table, echo=FALSE, results="asis"}
schedule <- schedule %>%  
    mutate(var_title = ifelse(!is.na(content),
                              glue('<span class="content-title">{title}</span>'),
                              glue('{title}'))) %>%
    mutate(var_deadline = ifelse(!is.na(deadline),
                                 glue('&emsp;&emsp;<small>(submit by {deadline})</small>'),
                                 glue(""))) %>%
    mutate(var_content = ifelse(!is.na(content),
                                glue('<a href="{content}.qmd"><i class="fa-solid fa-book-open-reader fa-lg"></i></a>'),
                                glue('<font color="#e9ecef"><i class="fa-solid fa-book-open-reader fa-lg"></i></font>'))) %>%
    mutate(var_tutorial = ifelse(!is.na(tutorial),
                                glue('<a href="{tutorial}"><i class="fa-solid fa-laptop-code fa-lg"></i></a>'),
                                glue('<font color="#e9ecef"><i class="fa-solid fa-laptop-code fa-lg"></i></font>'))) %>%
    mutate(var_assignment = ifelse(!is.na(assignment),
                                   glue('<a href="{assignment}"><i class="fa-solid fa-pen-ruler fa-lg"></i></a>'),
                                   glue('<font color="#e9ecef"><i class="fa-solid fa-pen-ruler fa-lg"></i></font>'))) %>%
    mutate(col_date = ifelse(is.na(date_end),
                             glue('<span>{format(date, "%B %e")}</span>'),
                             glue('<span>{format(date, "%B %e")}</span>–<span>{format(date_end, "%B %e")}</span>'))) %>%
    mutate(col_title = glue('<span class="content-title" style="font-weight: normal;">{var_title}</span>{var_deadline}')) %>%
    mutate(col_content = var_content,
           col_tutorial = var_tutorial,
           col_assignment = var_assignment) %>%
  select(Week = week, Date = col_date, Topic = col_title, Materials = col_content,
         Tutorial = col_tutorial, Assignment = col_assignment) 

week_count <- schedule |> count(Week)
week_index <- week_count |> pull(n) |> set_names(paste0("Week ", week_count$Week))
 
tbl <- schedule |>
  select(-Week) |>
  kbl(escap = FALSE, align = "llcccc", table.attr = 'class="schedule-table"') |>
  kable_styling() |>
  column_spec(1, width = "20%", extra_css = "padding-right: 20px;") |>  
  column_spec(2, width = "65%") |>  
  column_spec(c(3:5), width = "5%") |>
  pack_rows(index = week_index,
              label_row_css = "border-bottom: 2px solid #000000;")

  cat(tbl)
```
  
